---
title: "Bayesian Data Analysis - Assignment 9"
author: Anonymous
output: 
  pdf_document: 
    toc: yes
    toc_depth: 1
bibliography: bibliography.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/harvard-anglia-ruskin-university.csl
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries and data.
```{r, message=FALSE}
library(rstan)
library(ggplot2)
library(bayesplot)
library(aaltobda)
data("factory")
```

Set `rstan` options
```{r}
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

The course textbook @bayesian_data_analysis_2013.

# Subtitle
The hierarchical factory Stan model.
```{stan, output.var = "factory_stan"}
data {
  int<lower=0> N;  // number of observations per group
  int<lower=0> J;  // number of groups
  matrix[N, J] y;  // quality control measurement
}

parameters {
  vector[J] theta;  // mean
  real<lower=0> sigma;  // variance
  real mu;  // group mean
  real<lower=0> tau;  // group variance
}

model {
  for (j in 1:J) {
    theta[j] ~ normal(mu, tau);
  }
  for (j in 1:J){
    for (i in 1:N) {
      y[i, j] ~ normal(theta[j], sigma);
    }
  }
}

generated quantities {
  vector[J+1] y_pred;
  real theta_pred;
  theta_pred = normal_rng(mu, tau);
  for (j in 1:J) {
    y_pred[j] = normal_rng(theta[j], sigma);
  }
  y_pred[J+1] = normal_rng(theta_pred, sigma);
}
```

```{r, message = FALSE}
factory_data <- list(
  N = dim(factory)[1],
  J = dim(factory)[2],
  y = factory
)
fit_factory <- sampling(
  factory_stan, 
  data = factory_data, 
  control = list(max_treedepth = 15)
)
```

```{r}
print(fit_factory, pars = c("y_pred"), probs = c())
```

The utility function.
```{r}
utility <- function(draws, price = 200, cost = 106, quality_threshold = 85) {
  s = 0
  for (draw in draws) {
    if (draw < quality_threshold)
      s = s - cost
    else
      s = s - cost + price
  }
  s / length(draws)
}
```

Test case for utility function.
```{r}
y_pred <- c(123.80, 85.23, 70.16, 80.57, 84.91)
utility(y_pred)
```

The expectedutilities of machines 1 to 6.
```{r}
v = rstan::extract(fit_factory)
J = 6
utilities = rep(0, J)
for (j in 1:J) {
  utilities[j] = utility(v$y_pred[, j])
}
utilities
```

Machines ranked from worst to best. The indices denote the machine numbers.
```{r}
sort(utilities, index.return = TRUE)
``` 
Utility values denote the expected profit per product for a machine. For example, machine 1 has negative utility, and therefore is not profitable, the other machines are profitable.

The expected utility of 7th machine.
```{r}
utility(v$y_pred[, 7])
```

Company owner should buy the 7th machine because the prediction has positive utility.


# References
