---
title: "Bayesian Data Analysis - Assignment 2"
output: pdf_document
bibliography: bibliography.bib
---
## Inference for binomial proportion

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Regarding the equations, we refer to the course textbook, @bayesian_data_analysis_2013, chapters 2.1 - 2.4.

Let $\pi$ be the probability of a monitoring site having detectable blue-green algae levels.

In our Bayesian model:

- Likelihood $P(y\mid\pi)=\operatorname{Bin}(y\mid n,\pi)$ is **binomial distribution** where $n$ is the number of trials and $y$ the number of timers event $1$ happens.
- Prior $p(\pi)=\operatorname{Beta}(\alpha,\beta)$ is **beta distribution**.
- The resulting posterior $p(\pi\mid y)=\operatorname{Beta}(\alpha+y,\beta+n-y)$ is **beta distribution**.

---

Load the `algae` dataset. [@bda_course_aalto]
```{r}
library(aaltobda)
data("algae")
```

Algae dataset visualized.
```{r}
n = length(algae)
a = sum(algae == 0) 
b = sum(algae == 1)
barplot(c(a/n, b/n), ylim = c(0, 1.0), names.arg = c(0, 1))
```

```{r}
c(a, b, n)
```

## a)
Point estimate $E(\pi \mid y)=\frac{\alpha + y}{\beta + n + y}$
```{r}
beta_point_est <- function(prior_alpha, prior_beta, data) {
  n = length(data)
  y = sum(data == 1)
  (prior_alpha + y) / (prior_beta + n + y)
}
beta_point_est(prior_alpha = 2, prior_beta = 10, data = algae)
```

$90$% interval estimate
```{r}
beta_interval <- function(prior_alpha, prior_beta, data, prob) {
  n = length(data)
  y = sum(data == 1)
  offset = (1 - prob) / 2
  qbeta(c(offset, offset + prob), prior_alpha + y, prior_beta + n - y)
}
beta_interval(prior_alpha = 2, prior_beta = 10, data = algae, prob = 0.9)
```

## b)
$Pr(\pi < 0.2)$
```{r}
beta_low <- function(prior_alpha, prior_beta, data, pi_0) {
  n = length(data)
  y = sum(data == 1)
  pbeta(pi_0, prior_alpha + y, prior_beta + n - y)
}
beta_low(prior_alpha = 2, prior_beta = 10, data = algae, pi_0 = 0.2)
```

## c)
The random variables $\pi$ must be identically and independently distributed. This means that all lakes have an identical likelihood of having blue-green algae and if some lake has blue-green algae it doesn't affect the probability of other lakes having blue-green algae.

We also need to assume, that the prior distribution is conjugate, i.e. beta distribution in the case of binomial likelihood.


## d)
We can perform sensitivity analysis testing different value for `prior_alpha` and `prior_beta` parameters near the original values and comparing the results.
```{r}
for (prior_alpha in c(1, 2, 3)) {
  for (prior_beta in c(9, 10, 11)) {
    print(c(beta_interval(prior_alpha, prior_beta, data = algae, prob = 0.9),
          beta_point_est(prior_alpha, prior_beta, data = algae)))
  }
}
```
Looking at the 90% interval and the point estimates, we can see that the changes around the original parameter values don't produce very large changes. This indicates that model is not too sensitive.

## References
