---
title: "Bayesian Data Analysis - Assignment 1"
output: pdf_document
# bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1)
I rehearsed most of these with help from Wikipedia.

* **Probability** -- A measure of how likely an event is to occur.
* **Probability mass** -- Probability of a discrete random variable.
* **Probability mass function (pmf)** -- A function that gives the probability that a discrete random variable is exactly equal to some value.
* **Probability density function (pdf)** -- Specifies the probability of the continuous random variable falling within a particular range of values.
* **Probability distribution** -- Description of a random phenomenon in terms of the probabilities of events.
* **Discrete probability distribution** -- Probability distribution for discrete random variables.
* **Continuous probability distribution** -- Probability distribution for continuous random variables.
* **Cumulative distribution function (cdf)** -- Sum or integral from minus infinity to $x$ of a probability distribution.
* **Likelihood** -- Expresses how probable a given set of observations is for different values of statistical parameters.

## 2)
### a)
Plot of the Beta-distribution

```{r}
mu = 0.2
gamma_sq = 0.01
alpha = mu * (mu * (1 - mu) / gamma_sq - 1)
beta = alpha * (1 - mu) / mu
x = seq(0, 1, 0.01)
d = dbeta(x, alpha, beta)
plot(x, d)
```

### b)
Histogram of the random samples from Beta-distribution.

```{r}
samples = rbeta(1000, alpha, beta)
hist(samples)
```

Visually, the histogram looks similar to the Beta-distribution's density function as expected.

### c)
Mean and variance of the samples and their relative errors.

```{r}
m = mean(samples)
m
abs(mu - m) # Relative error
```

```{r}
v = var(samples)
v
abs(gamma_sq - v) # Relative error
```

Relative errors are small, thus the values roughly match the real values.

### d)
Estimation of the central 95%-interval

```{r}
quantile(samples, probs=c(0.95))
```


## 3)

To better understand how well the test works, we'll calculate the probabilities for true and false negatives and positives for a randomly sampled individual from the population. 

* True-negative -- No cancer and test shows no cancer
* False-positive -- No cancer, but test shows cancer
* True-positive -- Has cancer and test shows cancer
* False-negative -- Has cancer, but the test shows no cancer

The probabilities are

```{r}
0.999 * 0.96 # True negative
0.999 * 0.04 # False positive
0.001 * 0.98 # True positive
0.001 * 0.02 # False negative
```

As seen from the probabilities, the test has a high probability of showing a false positive. Using the test would lead to large numbers of unnecessary, potentially harmful, treatments.

## 4)
### a)
The probability of picking a red ball
```{r}
boxes <- matrix(
  c(2, 4, 1, 5, 1, 3), ncol=2,
  dimnames=list(c("A", "B", "C"), c("red", "white"))
)
p_red <- function(boxes) {
  0.4 * boxes["A", "red"] / sum(boxes["A",]) + 
  0.1 * boxes["B", "red"] / sum(boxes["B",]) + 
  0.5 * boxes["C", "red"] / sum(boxes["C",])
}
p_red(boxes)
```

### b)
```{r}
p_boxes <- function(boxes) {
  c(0.4 * boxes["A", "red"] / sum(boxes["A",]) / p_red(boxes),
  0.1 * boxes["B", "red"] / sum(boxes["B",]) / p_red(boxes), 
  0.5 * boxes["C", "red"] / sum(boxes["C",]) / p_red(boxes))
}
p_boxes(boxes)
```

The red ball has the highest probability of being picked from box C.

## 5)
The probability that fraternal twins are both boys is $1/4$ because the possible combinations are boy-boy, boy-girl, girl-boy, and girl-girl.

The probability that identical twins are both boys is $1/2$ because the possible combinations are boy-boy or girl-girl.

Therefore the probability that Elvis (male) had an identical brother (male) is given by the conditional probability:

```{r}
p_identical_twin <- function(fraternal_prob, identical_prob) {
  identical_prob / (fraternal_prob + identical_prob)
}
p_identical_twin(fraternal_prob = 1/4 * 1/150, identical_prob = 1/2 * 1/400)
```
