---
title: "Bayesian Data Analysis - Assignment 5"
author: Anonymous
output: 
  pdf_document: 
    toc: yes
    toc_depth: 1
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Generalized linear model: Bioassay with Metropolis
As for equations and algorithms used in this assignemnt, we refer to the course textbook, @bayesian_data_analysis_2013, chapters 3.7, 10, and 11.

Load libraries and data.
```{r}
library(mvtnorm)
library(rstan)
library(aaltobda)
data("bioassay")
```

## 1.
Define the prior and posterior densities as in assignment 4. The likelihood is given in the `aaltobda` package.
```{r}
p_log_prior <- function(alpha, beta) {
  mvtnorm::dmvnorm(c(alpha, beta), 
          mean = c(0, 10), 
          sigma = matrix(c(2^2, 10, 10, 10^2), nrow = 2), 
          log = TRUE)
}

p_log_posterior <- function(alpha, beta, x = bioassay$x, y = bioassay$y, n = bioassay$n) {
  p_log_prior(alpha, beta) + aaltobda::bioassaylp(alpha, beta, x, y, n)
}
```

### a)
The density ratio is defined as
$$
r = \frac{p(\theta^*\mid y)}{p(\theta^{t-1}\mid y)}.
$$
Using logarithmic densities we have
$$
r = \exp(\log p(\theta^*\mid y) - \log p(\theta^{t-1}\mid y)).
$$

```{r}
density_ratio <- function(
    alpha_propose, alpha_previous, beta_propose, beta_previous, 
    x = bioassay$x, y = bioassay$y, n = bioassay$n) {
  exp(p_log_posterior(alpha_propose, beta_propose) - 
      p_log_posterior(alpha_previous, beta_previous))
}
```


### b)
The Metropolis algorithm for the bioassay data. Starting points are drawn from a multivariate normal distribution. As the proposal distributions we use the following normal distributions
$$
\begin{aligned}
\alpha^* &\sim N(\alpha_{t-1}, \sigma_\alpha) \\
\beta^* &\sim N(\beta_{t-1}, \sigma_\beta).
\end{aligned}
$$
The standard deviations, $\sigma_\alpha$ and $\sigma_\beta$ can be set by the user.

```{r}
metropolis_bioassay <- function(sigma_alpha, sigma_beta, t_max) {
  # Set the starting point at t = 1.
  alpha <- numeric(t_max)
  beta <- numeric(t_max)
  values <- mvtnorm::rmvnorm(
    1, mean = c(0, 10), sigma = matrix(c(2^2, 10, 10, 10^2), nrow = 2))
  alpha[1] <- values[,1]
  beta[1] <- values[,2]
  
  # Iterate from t = 2 to t = t_max.
  for (t in 2:t_max) {
    # Samples the proposals.
    alpha_propose <- rnorm(1, alpha[t-1], sigma_alpha)
    beta_propose <- rnorm(1, beta[t-1], sigma_beta)
    
    # Calculate the ratio of the densities.
    r <- density_ratio(alpha_propose, alpha[t-1], beta_propose, beta[t-1])  
    
    # Set the values at time t.
    if (runif(1) <= min(r, 1)) {
      # `min(r, 1)` probability that the proposal is accepted.
      alpha[t] <- alpha_propose
      beta[t] <- beta_propose
    } else {
      alpha[t] <- alpha[t-1]
      beta[t] <- beta[t-1]
    }
  }
  
  # Return the values.
  list(alpha = alpha, beta = beta)
}
```


## 2.
We'll run two chains of the Metropolis algorithm using the parameters $\sigma_\alpha$, $\sigma_\beta$ and $t_{max}$ values below. We'll discard half of the iterations as warm up iterations.
```{r}
t_max = 2000
warm_up = floor(t_max / 2)
output1 = metropolis_bioassay(sigma_alpha = 1, sigma_beta = 5, t_max = t_max)
output2 = metropolis_bioassay(sigma_alpha = 1, sigma_beta = 5, t_max = t_max)
```

Plot of $\alpha$ in as a function of simulation time $t$.
```{r}
plot(1:t_max, output1$alpha, "l", xlab = "t", ylab = "alpha", col = "blue")
lines(1:t_max, output2$alpha, "l", xlab = "t", ylab = "alpha", col = "red")
```

Plot of $\beta$ in as a function of simulation time $t$.
```{r}
plot(1:t_max, output1$beta, "l", xlab = "t", ylab = "beta", col = "blue")
lines(1:t_max, output2$beta, "l", xlab = "t", ylab = "beta", col = "red")
```

Two chains, indicated by the colors blue and red, seem to convergence for both $\alpha$ and $\beta$ when visually assessing the convergence from the plots.

We can also verify this by computing the difference between the mean and variance of $\alpha$ and $\beta$ for both chains after the varm up iterations.
```{r}
abs(mean(output1$alpha[warm_up:t_max]) - mean(output2$alpha[warm_up:t_max]))
abs(var(output1$alpha[warm_up:t_max]) - var(output2$alpha[warm_up:t_max]))
```

```{r}
abs(mean(output1$beta[warm_up:t_max]) - mean(output2$beta[warm_up:t_max]))
abs(var(output1$beta[warm_up:t_max]) - var(output2$beta[warm_up:t_max]))
```
The values are small indicating convergence.


## 3.
The $\hat{R}$ for $\alpha$ and $\beta$ is calculated using $\hat{R}$ from the `rstan` library.

```{r}
rstan::Rhat(matrix(c(output1$alpha[warm_up:t_max], output2$alpha[warm_up:t_max]), ncol = 2))
```

```{r}
rstan::Rhat(matrix(c(output1$beta[warm_up:t_max], output2$beta[warm_up:t_max]), ncol = 2))
```

The values of $\hat{R}$ are less than $1.05$ (value from `rstan::Rhat` documentation) which indicates convergence for the chains.


## 4.
Scatter plot of $\alpha$ and $\beta$ for both chains.
```{r}
plot(output1$alpha[warm_up:t_max], output1$beta[warm_up:t_max], 
     "p", xlab = "alpha", ylab = "beta",
     ylim = c(-10, 40), xlim = c(-4, 10), 
     col = rgb(red = 0, green = 0, blue = 1, alpha = 1))
lines(output2$alpha[warm_up:t_max], output2$beta[warm_up:t_max], 
      "p", xlab = "alpha", ylab = "beta",
      ylim = c(-10, 40), xlim = c(-4, 10), 
      col = rgb(red = 1, green = 0, blue = 0, alpha = 1))
```

The resulting plot look similar to the Figure 3.3b in @bayesian_data_analysis_2013.


# Markmyassignment
```{r}
library(markmyassignment)
set_assignment(
  "https://github.com/avehtari/BDA_course_Aalto/blob/master/exercises/tests/ex5.yml")
mark_my_assignment()
```


# References
