---
title: "Bayesian Data Analysis - Assignment 8"
author: Anonymous
output: 
  pdf_document: 
    #toc: yes
    #toc_depth: 1
bibliography: bibliography.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/harvard-anglia-ruskin-university.csl
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries and data.
```{r, message=FALSE}
library(rstan)
library(loo)
library(ggplot2)
library(bayesplot)
library(aaltobda)
data("factory")
```

Set `rstan` options
```{r}
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

The course textbook @bayesian_data_analysis_2013.

# 1.
## Pooled Model
The pooled factory Stan model.
```{stan, output.var = "factory_stan1"}
data {
  int<lower=0> N;  // number of observations
  vector[N] y;  // quality control measurement
}

parameters {
  real theta;  // mean
  real<lower=0> sigma;  // variance
}

model {
  for (i in 1:N) {
    y[i] ~ normal(theta, sigma);
  }
}

generated quantities {
  vector[N] log_lik;
  for (i in 1:N) {
    log_lik[i] = normal_lpdf(y[i] | theta, sigma);
  }
}
```

```{r, message = FALSE}
factory_data1 <- list(
  N = prod(dim(factory)),
  y = unlist(factory)
)
fit_factory1 <- sampling(
  factory_stan1, 
  data = factory_data1, 
  control = list(max_treedepth = 15)
)
```


## Separate Model
The separate factory Stan model.
```{stan, output.var = "factory_stan2"}
data {
  int<lower=0> N;  // number of observations
  int<lower=0> J;  // number of groups
  matrix[N, J] y;  // quality control measurement
}

parameters {
  real theta[J];  // mean
  real<lower=0> sigma[J];  // variance
}

model {
  for (j in 1:J){
    for (i in 1:N) {
      y[i, j] ~ normal(theta[j], sigma[j]);
    }
  }
}

generated quantities {
  matrix[N, J] log_lik;
  for (j in 1:J) {
    for (i in 1:N) {
      log_lik[i, j] = normal_lpdf(y[i, j] | theta[j], sigma[j]);
    }
  }
}
```

```{r, message = FALSE}
factory_data2 <- list(
  N = dim(factory)[1],
  J = dim(factory)[2],
  y = factory
)
fit_factory2 <- sampling(
  factory_stan2, 
  data = factory_data2, 
  control = list(max_treedepth = 15)
)
```


## Hierarchical Model
The hierarchical factory Stan model.
```{stan, output.var = "factory_stan3"}
data {
  int<lower=0> N;  // number of observations per group
  int<lower=0> J;  // number of groups
  matrix[N, J] y;  // quality control measurement
}

parameters {
  vector[J] theta;  // mean
  real<lower=0> sigma;  // variance
  real mu;  // group mean
  real<lower=0> tau;  // group variance
}

model {
  for (j in 1:J) {
    theta[j] ~ normal(mu, tau);
  }
  for (j in 1:J){
    for (i in 1:N) {
      y[i, j] ~ normal(theta[j], sigma);
    }
  }
}

generated quantities {
  matrix[N, J] log_lik;
  for (j in 1:J) {
    for (i in 1:N) {
      log_lik[i, j] = normal_lpdf(y[i, j] | theta[j], sigma);
    }
  }
}
```

```{r, message = FALSE}
factory_data3 <- list(
  N = dim(factory)[1],
  J = dim(factory)[2],
  y = factory
)
fit_factory3 <- sampling(
  factory_stan3, 
  data = factory_data3, 
  control = list(max_treedepth = 15)
)
```


# 2. and 3.
The PSIS-LOO values and the $\hat{k}$-value, `p_loo`,  and the effective number of parameters $P_{eff}$, `elpd_loo`, for each model.

## Pooled model
```{r}
r1 <- loo::loo(fit_factory1)
print(r1)
```

```{r}
plot(r1)
```

## Separate model
```{r}
r2 = loo::loo(fit_factory2)
print(r2)
```

```{r}
plot(r2)
```

## Hierarchical model
```{r}
r3 = loo::loo(fit_factory3)
print(r3)
```

```{r}
plot(r3)
```


# 4.


# 5.


# References
