---
title: "Bayesian Data Analysis - Assignment 6"
author: Anonymous
output: 
  pdf_document: 
    toc: yes
    toc_depth: 1
bibliography: bibliography.bib
urlcolor: cyan
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Generalized linear model: Bioassay with Stan
As for equations and algorithms used in this assignemnt, we refer to the course textbook, @bayesian_data_analysis_2013, chapter 3.7. For features related to Stan we refer to the [Stan documentation](https://mc-stan.org/users/documentation/).

Load libraries and data.
```{r, message=FALSE}
library(ggplot2)
library(StanHeaders)
library(rstan)
library(aaltobda)
data("bioassay")
```

Set `rstan` options
```{r}
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```


## 1.
The model for the bioassay experiment in Stan syntax.
```{stan, output.var="bioassay_stan"}
data {
  int <lower=0> N;  // number of rows in the data set
  int <lower=0> n[N];
  vector[N] x;
  int y[N];
  vector[2] mu;  // mean 
  matrix[2, 2] sigma;  // variance (matrix)
}

parameters {
  vector[2] theta;  // (alpha, beta)
}

transformed parameters {
  vector[N] r;  // dose-response relation
  r = theta[1] + theta[2] * x;
}

model {
  theta ~ multi_normal(mu, sigma);  // prior 
  y ~ binomial_logit(n, r);  // likelihood
}

```

The input data for the stan model.
```{r}
data_bioassay = list(
  mu = c(0, 10),
  sigma = matrix(c(2^2, 10, 10, 10^2), nrow=2),
  N = dim(bioassay)[1],
  x = bioassay$x,
  y = bioassay$y,
  n = bioassay$n
)
```

Fit the Stan model to the data.
```{r}
fit <- sampling(bioassay_stan, data = data_bioassay)
```


## 2.
Rhat values close to $1$, under $1.05$, indicate convergence. 

Let print the values obtained from the fit.
```{r}
print(fit)
```

In the printout, `theta[1]` indicates $\alpha$ and `theta[2]` indicates $\beta$. Both have `Rhat` values close to $1$, which indicates that both have converged.


## 3.
Bayesplot is used for visulization of the results.
```{r, message=FALSE}
library(bayesplot)
```

Scatter plot of $\alpha$ (`theta[1]`) and $\beta$ (`theta[2]`).
```{r}
bayesplot::mcmc_scatter(fit, pars = c("theta[1]", "theta[2]"))
```
The resulting plot looks similar to the Figure 3.3b in @bayesian_data_analysis_2013.


# References
